---
title: "Lab 1 SDM Evaluate Models"
author: "Halina Do-Linh"
date: "1/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r}
# global knitr chunk options
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE)

# load packages
librarian::shelf(
  dismo, # species distribution modeling: maxent(), predict(), evaluate(), 
  dplyr, ggplot2, GGally, here, maptools, readr, 
  raster, readr, rsample, sf,
  usdm)  # uncertainty analysis for species distribution models: vifcor()
select = dplyr::select

# options
set.seed(42)
options(
  scipen = 999,
  readr.show_col_types = F)
ggplot2::theme_set(ggplot2::theme_light())

# paths
dir_data      <- here("data/sdm")
pts_geo       <- file.path(dir_data, "pts.geojson")
env_stack_grd <- file.path(dir_data, "env_stack.grd")
mdl_maxv_rds  <- file.path(dir_data, "mdl_maxent_vif.rds")

# read points of observation: presence (1) and absence (0)
pts <- read_sf(pts_geo)

# read raster stack of environment
env_stack <- raster::stack(env_stack_grd)
```


# 1.1 Split observations into training and testing 

```{r}
# create training set with 80% of full data
pts_split  <- rsample::initial_split(
  pts, prop = 0.8, strata = "present")
pts_train  <- rsample::training(pts_split)
pts_test   <- rsample::testing(pts_split)

pts_train_p <- pts_train %>% 
  filter(present == 1) %>% 
  as_Spatial()
pts_train_a <- pts_train %>% 
  filter(present == 0) %>% 
  as_Spatial()
```

# 2 Calibrate: Model Selection

```{r}
# show pairs plot before multicollinearity reduction with vifcor()
pairs(env_stack)
```

```{r}
# calculate variance inflation factor per predictor, a metric of multicollinearity between variables
vif(env_stack)
```

```{r}
# stepwise reduce predictors, based on a max correlation of 0.7 (max 1)
v <- vifcor(env_stack, th=0.7) 
v
```

**Halina Notes: One variable has a collinearity problem:**

**- `ER_tri`**

**After excluding `ER_tri`, the linear correlation coefficients ranges between:**

**- min correlation (`WC_bio2` ~ `ER_topoWet`): 0.009**
**- max correlation (`WC_alt` ~ `ER_topoWet`): 0.56**


```{r}
# reduce enviromental raster stack by 
env_stack_v <- usdm::exclude(env_stack, v)

# show pairs plot after multicollinearity reduction with vifcor()
pairs(env_stack_v)
```

```{r}
# fit a maximum entropy model
if (!file.exists(mdl_maxv_rds)){
  mdl_maxv <- maxent(env_stack_v, sf::as_Spatial(pts_train))
  readr::write_rds(mdl_maxv, mdl_maxv_rds)
}
mdl_maxv <- read_rds(mdl_maxv_rds)

# plot variable contributions per predictor
plot(mdl_maxv)
```

```{r}
# plot term plots
response(mdl_maxv)
```


```{r}
# predict
y_maxv <- predict(env_stack, mdl_maxv) #, ext=ext, progress='')

plot(y_maxv, main='Maxent, raw prediction')
data(wrld_simpl, package="maptools")
plot(wrld_simpl, add=TRUE, border='dark grey')
```


# 3 Evaluate: Model Performance

## 3.1 Area Under the Curve (AUC), Reciever Operator Characteristics (ROC) Curve and Confusion Matrix

```{r}
pts_test_p <- pts_test %>% 
  filter(present == 1) %>% 
  as_Spatial()
pts_test_a <- pts_test %>% 
  filter(present == 0) %>% 
  as_Spatial()

y_maxv <- predict(mdl_maxv, env_stack)
#plot(y_maxv)

e <- dismo::evaluate(
  p     = pts_test_p,
  a     = pts_test_a, 
  model = mdl_maxv,
  x     = env_stack)
e
```


```{r}
plot(e, 'ROC')
# add point to ROC plot
points(fpr, tpr, pch=23, bg="blue")

thr <- threshold(e)[['spec_sens']]
thr
```


```{r}
p_true <- na.omit(raster::extract(y_maxv, pts_test_p) >= thr)
a_true <- na.omit(raster::extract(y_maxv, pts_test_a) < thr)

# (t)rue/(f)alse (p)ositive/(n)egative rates
tpr <- sum(p_true)/length(p_true)
fnr <- sum(!p_true)/length(p_true)
fpr <- sum(!a_true)/length(a_true)
tnr <- sum(a_true)/length(a_true)

matrix(
  c(tpr, fnr,
    fpr, tnr), 
  nrow=2, dimnames = list(
    c("present_obs", "absent_obs"),
    c("present_pred", "absent_pred")))
```


```{r}
plot(y_maxv > thr)
```

