---
title: "Lab 1 SDM Logistic Regression"
author: "Halina Do-Linh"
date: "1/6/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Explore continued 

```{r}
librarian::shelf(
  DT, dplyr, dismo, GGally, here, readr, tidyr)
select <- dplyr::select # overwrite raster::select
options(readr.show_col_types = F)

dir_data    <- here("data/sdm")
pts_env_csv <- file.path(dir_data, "pts_env.csv")

pts_env <- read_csv(pts_env_csv)
nrow(pts_env)
```

```{r}
datatable(pts_env, rownames = F)
```

Creating pairs plots using `GGally` to show correlations between variables.
**Halina notes: consider dropping `ER_topoWet` and `ER_tri`**

```{r}
GGally::ggpairs(
  select(pts_env, -ID),
  aes(color = factor(present)))
```

# 2.1 Setup Data

```{r}
# setup model data
d <- pts_env %>% 
  select(-ID) %>%  # remove terms we don't want to model
  tidyr::drop_na() # drop the rows with NA values
nrow(d)
```

# 2.2 Linear Model

```{r}
# fit a linear model
mdl <- lm(present ~ ., data = d)
summary(mdl)
```

**Halina notes: `predict()` is creating a RasterLayer with a prediction based on a model, not 100% sure what the type argument is doing but I know we want the `y-predict` to be our response or y-values. But basically this is the part where we are running the logit regression so that our range is constrained between 0 and 1. We don't want predictions outside of this range.**

```{r}
y_predict <- predict(mdl, d, type="response")
y_true    <- d$present

range(y_predict)
```


```{r}
range(y_true)
```

The problem with these predictions is that it ranges outside the possible values of present 1 and absent 0. (Later we’ll deal with converting values within this range to either 1 or 0 by applying a cutoff value; i.e. any values > 0.5 become 1 and below become 0.)

**Halina note: also my adjusted r squared is pretty small?**

# 2.3 Generalized Linear Model

```{r}
# fit a generalized linear model with a binomial logit link function
mdl <- glm(present ~ ., family = binomial(link="logit"), data = d)
summary(mdl)
```


```{r}
y_predict <- predict(mdl, d, type="response")

range(y_predict)
```

Excellent, our response is now constrained between 0 and 1. Next, let’s look at the term plots to see the relationship between predictor and response.

```{r}
# show term plots
termplot(mdl, partial.resid = TRUE, se = TRUE, main = F)
```


# 2.4 Generalized Additive Model

```{r}
librarian::shelf(mgcv)

# fit a generalized additive model with smooth predictors
mdl <- mgcv::gam(
  formula = present ~ s(WC_alt) + s(WC_bio1) + 
    s(WC_bio2) + s(ER_tri) + s(ER_topoWet) + s(lon) + s(lat), 
  family = binomial, data = d)
summary(mdl)
```

**Halina note: well my adjusted r squared definitely increased!**


```{r}
# show term plots
plot(mdl, scale=0)
```

# 2.5 Maxtent (maximum entropy)

```{r}
# load extra packages
librarian::shelf(
  maptools, sf)

# show version of maxent
maxent()
```

```{r}
# get environmental rasters
# NOTE: the first part of Lab 1. SDM - Explore got updated to write this clipped environmental raster stack
env_stack_grd <- file.path(dir_data, "env_stack.grd")
env_stack <- stack(env_stack_grd)
plot(env_stack, nc=2)
```


```{r}
# get presence-only observation points (maxent extracts raster values for you)
obs_geo <- file.path(dir_data, "obs.geojson")
obs_sp <- read_sf(obs_geo) %>% 
  sf::as_Spatial() # maxent prefers sp::SpatialPoints over newer sf::sf class

# fit a maximum entropy model
mdl <- maxent(env_stack, obs_sp)
```


**Halina notes: I'm concerned about this warning - I thought I got rid of the NA's?**

```{r}
# plot variable contributions per predictor
plot(mdl)
```
**Halina Notes: interesting that ER_tri and ER_topoWet contribute the least to each predictor because they are also variables I would consider removing since their correlation number was greater than 0.7.**

```{r}
# plot term plots
response(mdl)
```

```{r}
# predict
y_predict <- predict(env_stack, mdl) #, ext=ext, progress='')

plot(y_predict, main='Maxent, raw prediction')
data(wrld_simpl, package="maptools")
plot(wrld_simpl, add=TRUE, border='dark grey')
```

